---
title: "Applied microeconometrics"
subtitle: "Week 3 - Discrete Choice"
author: "Josh Merfeld"
institute: "KDI School"
date: "2024-09-30"

date-format: long
format: 
  revealjs:
    self-contained: true
    slide-number: false
    progress: false
    theme: [serif, custom.scss]
    width: 1500
    height: 1500*(9/16)
    code-copy: true
    code-fold: show
    code-overflow: wrap
    highlight-style: github
execute:
  echo: true
  warnings: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = "png") # NOTE: switched to png instead of pdf to decrease size of the resulting pdf

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  #ifelse(options$size != "a", paste0("\n \\", "tiny","\n\n", x, "\n\n \\normalsize"), x)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})




library(tidyverse)
library(kableExtra)
library(fixest)
library(ggpubr)
library(RColorBrewer)
library(haven)
library(mfx)
library(nnet)
library(survival)
library(survminer)

df <- read_dta("week3files/data.dta")
data(iris)
multinomresults <- multinom(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, data = iris)

```



# Introduction

## What are we doing today?


- This week we will review methods for limited dependent variables
  - We will also discuss how to interpret the results


- Some of this will be review, but there will be some new stuff, too
  - For example, we will discuss poisson regression, which you might not have seen before




## What are we doing today?


- Much of the mathematical notation and theory comes from *Econometrics* by Bruce Hansen
  - I already discussed the older version of the textbook that is available for free online


- A small note:
  - You can use OLS for binary outcomes! This is actually pretty common in economics.
  - I'll discuss this more in a bit.
  - Other disciplines (e.g. public health) really like some of the new methods we will discuss today, so they are worth knowing.







# Binary Choice

## Binary choice


- Let's start with the simplest possibility: binary choice

- You can think of this as a No/Yes or False/True question, but we will generally refer to it as 0/1 choice
  - In programming, always remember that $FALSE=0$ and $TRUE=1$

- Focusing on the binary choice case will allow us to build intuition for the more general case of discrete choice
  - We will also be able to use the same data as we move to the more general case






## Dichotomous variables

- We will be thinking about this $Y\in\{0,1\}$
  - In other words, $Y$ is a dichotomous (dummy) variable that can only be equal to 0 or 1

- We are going to discuss methods to output *conditional probabilities*:
$$\begin{gather} \mathbb{P}\left[Y=1|X\right] \end{gather}$$






## The error term

- Consider the following model:
$$\begin{gather} Y = \beta X + \epsilon, \end{gather}$$
where $X$ can have any number of columns (variables), $k$.

- We already know that $\epsilon$ does not need to be normally distributed

- In fact, if $Y\in\{0,1\}$, then $\epsilon$ *will never be normally distributed*
  - Why?






## The error term

- $\epsilon$ has a two-point conditional distribution:
$$\begin{gather} \epsilon = \Biggl\{ \begin{array}{ll} 1 - P(X), & \text{with probability } P(X) \\ P(X), & \text{with probability } 1 - P(X) \end{array} \end{gather}$$

- $\epsilon$ is *heteroskedastic*:
$$\begin{gather} \text{Var}(\epsilon|X) = P(X)(1-P(X)) \end{gather}$$
- In fact, the variance of any dummy variable is $P(1-P)$, where $P$ is the probability of the dummy variable being equal to 1






## Scatterplots are pretty worthless!
:::: {.columns}

::: {.column width="50%"}
```{r scatter, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
# for reading in Stata data. 
# Install this using your console
library(haven) 
# read in data for the week:
df <- read_dta("data.dta")

# scatter of in_poverty on h_age:
ggplot(data = df) +
  geom_point(aes(x = h_age, y = in_poverty)) +
  labs(x = "Head age", y = "In poverty this month") +
  theme_bw() +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```
:::

::: {.column width="50%"}
```{r scatter2, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "100%", fig.align = "center"}
ggplot(data = df) +
  geom_point(aes(x = h_age, y = in_poverty)) +
  labs(x = "Head age", y = "In poverty this month") +
  theme_bw() +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```
:::
::::





## Let's plot out the conditional probabilities (means)
:::: {.columns}

::: {.column width="50%"}
```{r scatter3, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
# means by age
povmeans <- df %>%
            group_by(h_age) %>%
            summarize(mean = mean(in_poverty)) %>%
            ungroup

# scatter of in_poverty on h_age:
ggplot(data = povmeans) +
  geom_point(aes(x = h_age, y = mean)) +
  labs(x = "Head age", y = "P(poverty)") +
  theme_bw() +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```
:::

::: {.column width="50%"}
```{r scatter4, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "100%", fig.align = "center"}
# means by age
povmeans <- df %>%
            group_by(h_age) %>%
            summarize(mean = mean(in_poverty)) %>%
            ungroup

# scatter of in_poverty on h_age:
ggplot(data = povmeans) +
  geom_point(aes(x = h_age, y = mean)) +
  labs(x = "Head age", y = "P(poverty)", title = "Mean poverty by age of head") +
  theme_bw() +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```
:::
::::





## Linear probability model (LPM) 

- We can estimate this using OLS:
$$\begin{gather} Y = \beta X + \epsilon \end{gather}$$

```{r lpm, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "100%", fig.align = "center"}
f1 <- feols(in_poverty ~ h_male + h_age, data = df)
f2 <- feols(in_poverty ~ h_male + h_age, data = df, vcov = "HC1")
etable(f1, f2)
```







## Linear probability model (LPM) 

```{r lpm2, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "100%", fig.align = "center"}
summary(feols(in_poverty ~ h_male + h_age, data = df, vcov = "HC1"))
```


- The interpretation of these coefficients is pretty straightforward:
  - It is the change in the probability of being in poverty for a one-unit change in the variable of interest
  - Male-headed households are 6.1 *percentage points* less likely to be poor than female-headed households, controlling for age.
  - Each addition year of age *increases* the probability of being in poverty by 0.005 *percentage points*, controlling for gender of the head.






## Linear probability model (LPM) 

- Sometimes people motivate other estimation methods based on heteroskedasticity
  - But we can easily correct for this using robust standard errors (HC1 in `feols`)

- There are two other problems with LPM, though:

. . .

- The predicted values can be outside of the 0-1 range
  - Is this a problem? Maybe. Maybe not. It depends on what you're doing. \pause
- Constant effects throughout the probability distribution

. . .

  - Is this realistic? If we think someone has a 95 percent probability of being poor, do we think the percentage point change would be the same for changing a variable relative to someone with a 50 percent probability of being poor?







## Another option


- We can think of this as a *latent variable* model:
$$\begin{gather} Y^* = \beta X + \epsilon \\
                \epsilon\sim G(\epsilon) \\
                Y = \mathbbm{1}(Y^*>0) = \Biggl\{ \begin{array}{ll} 1, & \text{if } Y^*>0 \\ 0, & \text{otherwise} \end{array} \end{gather}$$
where $Y^*$ is the latent variable, $G(\cdot)$ is the distribution of $\epsilon$, and $\mathbbm{1}(\cdot)$ is the indicator function.

- One way to think about this is that $y^*$ is utility, but we only observe whether the choice increases utility ($y=1$) or doesn't ($y=0$).








## Let's give this a bit more structure


- Note that $Y=1$ iff $Y^*>0$, which is the same as saying $\beta X + \epsilon > 0$

- The response probability is then given by the CDF of $\epsilon$ evaluated at $-\beta X$:
$$\begin{gather} \mathbb{P}\left[Y=1|X\right] = \mathbb{P}\left[\epsilon > -\beta X\right] = 1 - G(-\beta X) = G(\beta X) \end{gather}$$

- Note that CDFs (cumulative distribution functions) give us probabilities of being *less than or equal to* a given value
  - The last equality holds because $G(\cdot)$ will always be *symmetric around 0* here
  - That value here is $\beta X$







## CDF examples of the logistic (sigmoid) function - Wikipedia

![](week3files/cdfwiki)







## The link function

- The function $G(\cdot)$ is called the *link function* and plays an important role here

- Two common link functions are the *logit* and *probit* link functions:
  - They are defined as follows:
  - Logit: $G(\epsilon) = \frac{e^{\epsilon}}{1+e^{\epsilon}} = \frac{1}{1+e^{-\epsilon}}$
  - Probit: $G(\epsilon) = \Phi(\epsilon)$, where $\Phi(\cdot)$ is the CDF of the standard normal distribution

- We will discuss these in more detail in a bit
 







## Likelihood

- Likelihood: the joint probability of the data evaluated with the sample, as a function of the parameters
  - What?

- Let's start with probit, which uses a normal distribution. Here is the conditional density of $Y$ given $X$ under this assumption:
$$\begin{gather} f(Y|X) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{1}{2\sigma^2}(Y-\beta X)^2\right) \end{gather}$$








## Likelihood

$$\begin{gather} f(Y|X) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{1}{2\sigma^2}(y-\beta X)^2\right) \end{gather}$$

- In this case, what is the probability we *observe our sample given the values of $\beta$ and $\sigma$*?
$$\begin{align} f(y_1,\ldots,y_n | x_1,\ldots, x_n) &= \prod_{i=1}^n f(y_i | x_i) \\
                                                  &= \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{1}{2\sigma^2}(y_i-\beta x_i)^2\right) \\
                                                  &= \frac{1}{(2\pi\sigma^2)^{n/2}}\exp\left(-\frac{1}{2\sigma^2}\prod_{i=1}^n(y_i-\beta x_i)^2\right) \\
                                                  &= L_n(\beta, \sigma^2) \end{align}$$








## Likelihood

$$\begin{gather} f(y_1,\ldots,y_n | x_1,\ldots, x_n) = L_n(\beta, \sigma^2) \end{gather}$$

- This is the *likelihood function* 
  
  - Note that it is a function of the parameters, $\beta$ and $\sigma^2$

- The properties of logs make this easier to work with:
$$\begin{gather} \ell_n(\beta, \sigma^2) = \log(L_n(\beta, \sigma^2)) = -\frac{n}{2}\log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n(y_i-\beta x_i)^2 \end{gather}$$
- This is the *log likelihood function* 
  
  - It is of course also a function of the parameters, $\beta$ and $\sigma^2$








## Maximum likelihood estimation


$$\begin{gather} \ell_n(\beta, \sigma^2) = \log(L_n(\beta, \sigma^2)) = -\frac{n}{2}\log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n(y_i-\beta x_i)^2 \end{gather}$$


- We have our log likelihood function, which is a function of the parameters, $\beta$ and $\sigma^2$








## Maximum likelihood estimation
  

- What we want to do is find the values of $\beta$ and $\sigma^2$ that *maximize* this function
  - In other words, the values that make our sample the most likely to have been observed, or the biggest probability of observing our sample
  

- This is called *maximum likelihood estimation*

$$\begin{gather} \left(\hat{\beta}, \hat{\sigma}^2\right) = \underset{\beta\in\mathbb{R}^k,\sigma^2>0}{\text{argmax}}\;\ell_n(\beta, \sigma^2) \end{gather}$$
where $k$ is the number of variables (coefficients), including the intercept.








## Maximum likelihood estimation

- A simple example is a coin flip

- Let's say we flip a coin. If it's a fair coin, what is the probability of obtaining heads?

. . .

    - 50% or 0.5 (we generally work with the proportion 0.5, and not the percent)

. . .

- What is the probability of obtaining heads twice in a row?

. . .

    - 0.5 * 0.5 = 0.25

- Three times in a row?

. . .

    - 0.5 * 0.5 * 0.5 = 0.125








## Maximum likelihood estimation

- Say we flip the coin a bunch of times
  - For argument's sake, let's say we flip it 100 times and obtain 60 heads

- If we know nothing about whether the coin is actually fair, what is the *most likely distribution* that would give us 60 heads and 40 tails? \pause
  - It's a distribution in which the probability of heads is 0.6! \pause

- This is like maximum likelihood estimation. We are trying to find the parameters that makes our sample the most likely to have been observed.
  - In this case, the parameter would be the true mean of the distribution of the coin (where heads = 1 and tails = 0).
  - We could then of course test whether this is significantly different from 0.5, which might be our null hypothesis.









## Generalizing maximum likelihood estimation

$$\begin{gather} \left(\hat{\beta}, \hat{\sigma}^2\right) = \underset{\beta\in\mathbb{R}^k,\sigma^2>0}{\text{argmax}}\;\ell_n(\beta, \sigma^2) \end{gather}$$

- MLE is generally always estimated using numerical optimization
  - We will not discuss the details of this here
  - The basic reason is that most likelihood functions are not easy to maximize analytically (i.e. they have no closed-form solution)


- In the case of the normal regression model, however, there is a closed-form solution
  - And this is the same closed-form solution as OLS!









## Logit for binary choice

- Let's return to our binary choice model.
  - Regardless of how you estimate it, the probability mass function for $Y$ is:

$$\begin{gather} \pi(y) = p^y(1-p)^{1-y}, \end{gather}$$
where $p$ is the probability of $Y=1$, or the mean. Remember that $Y\in\{0,1\}$; i.e., it can only equal 0 or 1.

- Let's bring our link function back into it. The *conditional* probability is:
$$\begin{gather} \pi(Y|X)=G(\beta X)^Y(1-G(\beta X))^{1-Y}=G(\beta X)^Y(G(-\beta X))^{1-Y}=G(\beta Z),   \end{gather}$$

where $Z = \Biggl\{ \begin{array}{ll} X, & \text{if } Y=1 \\ -X, & \text{if } Y=0 \end{array}$








## Logit for binary choice

$$\begin{gather} \pi(Y|X)=G(\beta Z),   \end{gather}$$

- Taking logs (because they're easy to work with), we get the log likelihood function:
$$\begin{gather} \ell_n(\beta) = \sum_{i=1}^n \log G(\beta Z) \end{gather}$$


- This is the same as the log likelihood function for probit, except that the link function is different.








## Logit for binary choice

- Again, we want to find the values of $\beta$ (and $\sigma$, which will show up in the link function) that maximize this function.
$$\begin{gather} \left(\hat{\beta}, \hat{\sigma}^2\right) = \underset{\beta\in\mathbb{R}^k,\sigma^2>0}{\text{argmax}}\;\ell_n(\beta, \sigma^2) \end{gather}$$

- Something interesting is that in practice, we don't numerically maximize...
  - Instead, we *minimize* the *negative* of the log likelihood function!

$$\begin{gather} \left(\hat{\beta}, \hat{\sigma}^2\right) = \underset{\beta\in\mathbb{R}^k,\sigma^2>0}{\text{argmin}}\;-\ell_n(\beta, \sigma^2) \end{gather}$$








## An example in R - Household variables and poverty using glm()
```{r mle1, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
summary(glm(in_poverty ~ h_male, data = df, family = binomial(link = "logit")))
```







## Interpreting logit output
```{r mle2, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
summary(glm(in_poverty ~ h_male, data = df, family = binomial(link = "logit")))$coefficients
```


- How do we interpret these coefficients?
  - The coefficients are "log odds"

- For male household heads, the log odds of being in poverty is 0.280 *lower* than that for female household heads
  - What?








## Log odds

- What are log odds?
  - Let's start with odds:

$$\begin{gather} \text{odds} = \frac{p}{1-p}, \end{gather}$$

where $p$ is the probability of $y = 1$ (being poor in this case).

- Log odds?

$$\begin{gather} \text{log odds} = \log\left(\frac{p}{1-p}\right) \end{gather}$$







## Log odds and logit

- Logit regression is basically estimating:

$$\begin{gather} \log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X_1 + \ldots + \beta_k X_k \end{gather}$$









## The intercept is the log odds of being in poverty for female households
```{r mle3, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
summary(glm(in_poverty ~ h_male, data = df, family = binomial(link = "logit")))$coefficients
```

$$\begin{gather} \log\left(\frac{p}{1-p}\right) = -0.6196447 \\
                   \left(\frac{p}{1-p}\right) = exp(-0.6196447) \\
                   \left(\frac{p}{1-p}\right) \approx 0.538  \\ 
                   p = 0.538-0.538p \\
                   1.538p = 0.538 \\
                        p\approx 0.350  \end{gather}$$

What is the actual mean for female headed households? `r format(mean(df$in_poverty[df$h_male==0]), digits = 3)`








## The coefficient?

$$\begin{gather} \log\left(\frac{p_m}{1-p_m}\right) - \log\left(\frac{p_f}{1-p_f}\right) = -0.2795628 \\
                \log\left(\frac{\frac{p_m}{1-p_m}}{\frac{p_f}{1-p_f}}\right) = -0.2795628 \\
                \left(\frac{\frac{p_m}{1-p_m}}{\frac{p_f}{1-p_f}}\right) = exp(-0.2795628) \\
                \left(\frac{\frac{p_m}{1-p_m}}{\frac{p_f}{1-p_f}}\right) = 0.7561142 \end{gather}$$


- In the last line, this is referred to as an *odds ratio*. 
  - It's less than one, which means male-headed households are *less likely* to be in poverty.
  - Their *odds* of being in poverty are around 24% lower.
 







## The coefficient?

- Mean for female-headed households: `r format(mean(df$in_poverty[df$h_male==0]), digits = 3)`
  - odds ($\frac{p_f}{1-p_f}$): `r format(mean(df$in_poverty[df$h_male==0])/(1-mean(df$in_poverty[df$h_male==0])), digits = 3)`

- Mean for male-headed households: `r format(mean(df$in_poverty[df$h_male==1]), digits = 3)`
  - odds $\left(\frac{p_m}{1-p_m}\right)$: `r format(mean(df$in_poverty[df$h_male==1])/(1-mean(df$in_poverty[df$h_male==1])), digits = 3)`

- Odds ratio ($\frac{\frac{p_m}{1-p_m}}{\frac{p_f}{1-p_f}}$): `r format((mean(df$in_poverty[df$h_male==1])/(1-mean(df$in_poverty[df$h_male==1])))/(mean(df$in_poverty[df$h_male==0])/(1-mean(df$in_poverty[df$h_male==0]))), digits = 3)`
 
 - Exponentiating shows us the odds ratio!








## Marginal effects

- We can also calculate marginal effects
  - These are the change in the probability of being in poverty for a one-unit change in the variable of interest
  - An important note is that this *depends on where you are located in the distribution*

- We just calculated the means, so with only the binary independent variable, we know that the marginal effect is:
  - `r format(mean(df$in_poverty[df$h_male==1]) - mean(df$in_poverty[df$h_male==0]), digits = 3)`

- We will use the "mfx" package for this, so please install it in the console.











## Marginal effects
```{r mle4, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
logitmfx(in_poverty ~ h_male, data = df)

# for binary outcomes, it shows the change from 0 to 1!
```










## logit with more coefficients
```{r mle5, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
logitmfx(in_poverty ~ h_male + h_age, data = df)

# for binary outcomes, it shows the change from 0 to 1!
# for continuous variables, it's the derivative (i.e. instantaneous change)!
# By default, it calculates these by holding variables AT THEIR MEANS
```







## Probit
```{r probit1, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
summary(glm(in_poverty ~ h_male, data = df, family = binomial(link = "probit")))$coefficients
```


- What about probit coefficients?
  - These relate to the *CDF of the standard normal distribution*



- The intercept is the mean for female-headed households 









## Standard normal CDF
- The intercept is -0.3856924
  - The mean poverty for female-headed households is `r format(mean(df$in_poverty[df$h_male==0]), digits = 3)`
- Here's the CDF for the standard normal distribution with the intercept:

```{r probit2, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
ggplot() +
  stat_function(data = data.frame(x = c(-3, 3)), aes(x = x), fun = pnorm) +
  theme_bw() +
  labs(x = "x", y = "CDF(x)") +
  geom_vline(xintercept = -0.3856924, linetype = "dotted", color = "blue") +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```







## Standard normal CDF
- The intercept is -0.3856924
  - The mean poverty for female-headed households is `r format(mean(df$in_poverty[df$h_male==0]), digits = 3)`
- Here's the CDF for the standard normal distribution with BOTH:

```{r probit3, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
ggplot() +
  stat_function(data = data.frame(x = c(-3, 3)), aes(x = x), fun = pnorm) +
  theme_bw() +
  labs(x = "x", y = "CDF(x)") +
  geom_vline(xintercept = -0.3856924, linetype = "dotted", color = "blue") +
  geom_hline(yintercept = mean(df$in_poverty[df$h_male==0]), linetype = "dotted", color = "blue") +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```







## Now let's look at the coefficient on h_male
```{r probit4, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "65%", fig.align = "center"}
summary(glm(in_poverty ~ h_male, data = df, family = binomial(link = "probit")))$coefficients
```








## Standard normal CDF
- The intercept is -0.3038412 and the coefficient is -0.1699918
- Here's the CDF for the standard normal distribution with BOTH:

```{r probit5, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
ggplot() +
  stat_function(data = data.frame(x = c(-3, 3)), aes(x = x), fun = pnorm) +
  theme_bw() +
  labs(x = "x", y = "CDF(x)") +
  geom_vline(xintercept = -0.3038412, linetype = "dotted", color = "blue") +
  geom_vline(xintercept = -0.3038412-0.2540258, linetype = "dotted", color = "orange") +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```








## Standard normal CDF
- The intercept is -0.3856924 and the coefficient is -0.1699918
- What's the change in PROBABILITY? $\text{mean(male)} - \text{mean(female)}$ or `r format(mean(df$in_poverty[df$h_male==1]) - mean(df$in_poverty[df$h_male==0]), digits = 3)`

```{r probit6, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
ggplot() +
  stat_function(data = data.frame(x = c(-3, 3)), aes(x = x), fun = pnorm) +
  theme_bw() +
  labs(x = "x", y = "CDF(x)") +
  geom_vline(xintercept = -0.3856924, linetype = "dotted", color = "blue") +
  geom_vline(xintercept = -0.3856924-0.1699918, linetype = "dotted", color = "orange") +
  geom_hline(yintercept = mean(df$in_poverty[df$h_male==0]), linetype = "dotted", color = "blue") +
  geom_hline(yintercept = mean(df$in_poverty[df$h_male==1]), linetype = "dotted", color = "orange") +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```








## Marginal effects
- The intercept is -0.3856924 and the coefficient is -0.1699918
- What's the change in PROBABILITY? $\text{mean(male)} - \text{mean(female)}$ or `r format(mean(df$in_poverty[df$h_male==1]) - mean(df$in_poverty[df$h_male==0]), digits = 3)`


```{r probit7, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
probitmfx(in_poverty ~ h_male, data = df)
```








## Probit and marginal effects


```{r probit8, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# change in z-scores
summary(glm(in_poverty ~ h_male + h_age, data = df, family = binomial(link = "probit")))$coefficients
```


```{r probit9, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# change in probability, holding other variables at their means
probitmfx(in_poverty ~ h_male + h_age, data = df)
```






# Multiple discrete choice

## What if the outcome has more than two categories?

- Many outcomes are not 0/1.

- We can think of outcomes with discrete categories, but more than two:
  - Religion
  - Political party
  - Opinion on a likert scale (e.g. strongly agree, agree, neutral, disagree, strongly disagree)
  - Months in poverty

- There's a key difference between the first two and the last two:
  - The first two are *unordered*
  - The last two are *ordered*







## Months in poverty - distribution
```{r ologit1, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
ggplot() +
    geom_histogram(data = df, aes(x = months_in_poverty), binwidth = 1) + 
    theme_bw() +
    labs(x = "Months in poverty", y = "Count") +
    scale_x_continuous(breaks = seq(0, 12, 1)) +
    geom_vline(xintercept = mean(df$months_in_poverty), linetype = "dotted", color = "blue") +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```






## Ordered logistic regression with the polr function

```{r ologit2, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# note that the outcome must be a FACTOR variable for polr
summary(polr(as_factor(months_in_poverty) ~ h_male + h_age, data = df, Hess = TRUE))
```








## Ordered data: ordinal logit/probit

- When we have ordered discrete variable, we can use an ordered logit or probit model
  - These are also called *ordinal* logit/probit models

- The idea is that we have a latent variable, $Y^*$, that is continuous
  - We observe $Y$ as a discrete variable, but it is *ordered*
  - We can think of $Y$ as being *binned* into categories

$$\begin{gather} Y = \Biggl\{ \begin{array}{ll} 1, & \text{if } Y^*\in(-\infty,\theta_1] \\ 2, & \text{if } Y^*\in(\theta_1,\theta_2] \\ \vdots & \vdots \\ J, & \text{if } Y^*\in(\theta_{J-1},\infty) \end{array} \end{gather}$$





## The interpretation?

```{r ologit3, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# note that the outcome must be a FACTOR variable for polr
summary(polr(as_factor(months_in_poverty) ~ h_male + h_age, data = df, Hess = TRUE))
```
- The interpretation is similar to logit: a change in the log-odds of being in a higher level of months in poverty!





## Understanding fit with MLE

- There is no r-squared in MLE
  - It is not a true r-squared *because there is no sense of "mean" with discrete data, especially unordered data*
 
- We can use the log likelihood function to compare models
  - The log likelihood function is a function of the parameters, $\beta$ and $\sigma^2$
  - The higher the log likelihood, the better the fit

- We can also use the *Akaike Information Criterion* (AIC) and *Bayesian Information Criterion* (BIC)
  - These are functions of the log likelihood function and the number of parameters
  - The lower the AIC/BIC, the better the fit

- These are best thought of as useful for comparing across models
  - Difficult to interpret them on their own





## AIC

- AIC is defined as follows:
  - $k$ is the number of parameters in the model
  - $L$ is the log likelihood function
  - AIC: $2k - 2\log(L)$

```{r aic, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# save our model
results <- glm(in_poverty ~ h_male + h_age, data = df, family = binomial(link = "probit"))

# log likelihood
logLik(results)

# aic from temp
results$aic

# Calculate AIC by hand:
2*3 - 2*(-2788.727)
```





## BIC

- BIC is defined as follows:
  - $k$ is the number of parameters in the model
  - $L$ is the log likelihood function
  - $n$ is the number of observations
  - BIC: $k\log(n) - 2\log(L)$

```{r bic, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# save our model
results <- glm(in_poverty ~ h_male + h_age, data = df, family = binomial(link = "probit"))

# log likelihood
logLik(results)
nrow(df)

# Calculate BIC by hand:
3*log(4609) - 2*(-2788.727)
```







## Pseudo r-squared

- We can also calculate a pseudo r-squared
  - This is a measure of the change in the log likelihood function relative to the null model (no coefficients except intercepts)

```{r pseudor2, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# null model
logLik(glm(in_poverty ~ 1, data = df, family = binomial(link = "probit")))
# our model
logLik(glm(in_poverty ~ h_male + h_age, data = df, family = binomial(link = "probit")))
# pseudo r-squared: 1 - (log likelihood of model / log likelihood of null model)
1 - (-2788.727/-2791.606)
# Same thing: (null - model) / null
(-2791.606 - -2788.727)/-2791.606
```






## Multinomial probit/logit

- Ordered logit/probit is used when the outcome is *ordered*

- But what if it's not, like trying to predict what the species of a flower is?
  - Let's use a built-in dataset in R called "iris" to look at this:

```{r iris1, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
data(iris)
head(iris)
```






## Multinomial probit/logit

```{r iris2, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
data(iris)
table(iris$Species)
colnames(iris)
```

- Let's use sepal/pedal length/width to predict the species






## Multinomial probit/logit

```{r iris3, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
data(iris)
library(nnet)
multinomresults <- multinom(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, data = iris)
```

```{r iris4, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
summary(multinomresults)$coefficients
```


- Note how setosa isn't there... it's the omitted category
  - We can interpret the coefficients as the log odds of being in a particular category relative to setosa (the omitted category)








# Extensions

## Extensions

- GLM is a very general framework
  - We can use it for other distributions, too

- Let's look at a poisson distribution
  - The poisson distribution is often used for count data
  - It has assumptions (mean = variance), but violation isn't a big deal!
 




## The poisson distribution

- The poisson distribution is defined as follows:
$$\begin{gather} f(y; \lambda) = \frac{\lambda^ye^{-\lambda}}{y!} \end{gather}$$

- Note the $e$: this is going to lead to a nice log interpretation
- $y$ is the number of occurrences of the variable (in our example, it will be number of months in poverty)

- As mentioned, one implication of this distribution is:
$$\begin{gather} E(y) = Var(y) = \lambda \end{gather}$$
i.e. the mean of $y$ equals its variance. But we can work around this if it's false (which it probably is).





## Possion, months in poverty (with feols - feglm)

```{r poisson1, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# could save results. Not going to here... just display them
summary(feglm(months_in_poverty ~ h_male + h_age, data = df, family = "poisson"))
# notice the difference in standard errors if we use HC1 (which we want to here because of overdispersion)
summary(feglm(months_in_poverty ~ h_male + h_age, data = df, family = "poisson", vcov = "HC1"))
```





## Interpreting poisson coefficients

```{r poisson2, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
summary(feglm(months_in_poverty ~ h_male, data = df, family = "poisson", vcov = "HC1"))$coefficients
```


- How do we interpret poisson coefficients?
  - They are the *log* of the *rate* of the outcome
  - i.e. the log of the number of months in poverty

- The log count for male-headed households is around 0.28 *lower* than for female-headed households





## The intercept is log count for female households

```{r poisson3, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
summary(feglm(months_in_poverty ~ h_male, data = df, family = "poisson", vcov = "HC1"))$coefficients
# intercept is log count, so exponentiate for levels
exp(1.5189645)
# what is the mean for female-headed households?
mean(df$months_in_poverty[df$h_male==0])
# intercept plus coefficient for male-headed households
exp(1.5189645 - 0.2772304)
# mean for male-headed households?
mean(df$months_in_poverty[df$h_male==1])
```





## Quasi-poisson

- The poisson distribution has a mean = variance assumption
  - This is often violated
  - We can use quasi-poisson instead

- The quasi-poisson is the same as poisson, but the variance is estimated from the data
  - With feglm, it estimates the variance as a *linear function of the mean*

- Small note: if you use glm, you can use the "quasipoisson" family
  - This is more similar to poisson with vcov = "HC1"!
    - If you use "HC1" in both, you'll get identical results.
  - This is only about the structure of the error term. Not the coefficients.





## Possion vs quasi-poisson (note the similar standard errors with HC1 for poisson)

```{r poisson4, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
# Poisson with HC1
summary(feglm(months_in_poverty ~ h_male + h_age, data = df, family = "poisson", vcov = "HC1"))
# quasipoisson
summary(feglm(months_in_poverty ~ h_male + h_age, data = df, family = "quasipoisson"))
```





## Doesn't have to be integers!

- You can use poisson for non-integer outcomes, too!
  - It's just a distribution
  - It's often used for integer outcomes because it's a count distribution

- Jeff Wooldridge is a huge proponent of using (quasi) poisson

- Two really nice things: it is robust and has an easy interpretation
 



# Survival models

## Hazard/survival models

- These models are used for *duration* data
  - i.e. time until an event occurs or a state changes
  - e.g. time until death after being diagnosed with cancer, time until a person leaves poverty, time until a person gets a job, until contracting a disease, etc.

- You need a very specific kind of data for this...
  - We want to know what kinds of variables predict the occurence of some event (e.g. death).

- **Warning:** I am not an expert on survival models. I will give you a brief overview, but you should consult a textbook for more information.






## Pfizer-BioNTech COVID-19 vaccine

![](week3files/covidhazard)






## Survival models

- Let's use the package `survival` in R.
  - It has a dataset set up for us, called diabetic

```{r surv1, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
library(survival)
# dataset with "high-risk" diabetic retinopathy patients
head(diabetic)
```


- Note: this is a *panel* dataset
  - Each row is a patient (id)
  - Each patient has multiple observations (time)
  - Interested in loss of sight (status = 1)






## Survival function
 
- Let's first look at the survival function
  - This is the probability of surviving past a certain time
  - We're going to use the Kaplan-Meier estimator

```{r surv2, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
KM <- survfit(Surv(time = time, event = status) ~ 1, data = diabetic)
# note that Surv() is necessary here. It's a function that creates a survival object.
# The ~ 1 means this is for EVERYONE.
```






## Survival curve

```{r surv4, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
KM <- survfit(Surv(time = time, event = status) ~ 1, data = diabetic)
plot(KM, ylab = "Survival probability", xlab = "Time (months)")
```







## Comparing survival curves based on treatment

- Base R has ugly plots. We can use `survminer` to make it work with `ggplot2`.

```{r surv5, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
library(survminer)
KM <- survfit(Surv(time = time, event = status) ~ 1, data = diabetic)
ggsurvplot(KM)$plot +
  labs(y = "Survival probability", x = "Time (months)") +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```






## Comparing survival curves based on treatment

```{r surv6, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
KM <- survfit(Surv(time = time, event = status) ~ trt, data = diabetic)
ggsurvplot(KM)$plot +
  labs(y = "Survival probability", x = "Time (months)") +
  scale_color_discrete("Treatment:", labels = c("No", "Yes")) +
  theme(legend.position = c(0.1, 0.2)) +
  theme(plot.background = element_rect(fill = "#f0f1eb", color = "#f0f1eb"))
```






## Comparing survival curves based on treatment, empirically

```{r surv7, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
survdiff(Surv(time = time, event = status) ~ trt, data = diabetic)

# changes the weighting (more weight on earlier time points); doesn't matter here! Huge differences.
survdiff(Surv(time = time, event = status) ~ trt, data = diabetic, rho = 1)
```






## Adding more covariates

- Plotting two survival curves with a simply treatment dummy is straightforward.
  - But what if we want to add more covariates?
  - For example, the `diabetic` dataset has age at diagnosis and which eye the problem is. Does this matter?

- We can use a Cox proportional hazards model to do this.
  - This is a semi-parametric model that is very popular in survival analysis.
  - It is a *proportional hazards* model, which means that the hazard ratio is constant over time.
  - Its nature means we do not estimate the baseline hazard function.
    - Instead, we compare across variables
  





## Adding more covariates

```{r surv8, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
coxph(Surv(time = time, event = status) ~ age + as_factor(eye) + trt, data = diabetic)
```


- Note that treatment is randomized, so we *shouldn't* see large changes in the coefficient on treatment when we add covariates.
  - More on this next week!






## Some warnings

- Note that all these methods have assumptions that can sometimes be important

- Given our time, I am purposefully just showing you the basics
  - If any of these specific methods interest you, I suggest doing more in-depth readings. I can provide some suggestions.






## Articles for next week (check syllabus for textbook readings)

- Deserranno, E., Stryjan, M., & Sulaiman, M. (2019). Leader selection and service delivery in community groups: Experimental evidence from Uganda. *American Economic Journal: Applied Economics*, 11(4), 240-267.

- Fischer, T., Frölich, M., & Landmann, A. (2023). Adverse Selection in Low-Income Health Insurance Markets: Evidence from an RCT in Pakistan. *American Economic Journal: Applied Economics*, 15(3), 313-340.






