---
title:  |
    | Microeconometrics
    | Week 1 - Introduction to R
author:
   |
    | Joshua D. Merfeld
    | KDI School
date: "`r Sys.Date()`"

# Output type and options (no TOC and yes fig captions)
output: 
  beamer_presentation:
    theme: Montpellier
classoption: "aspectratio=169"

# This includes latex arguments
header-includes:
  - \AtBeginDocument{\title[Week 1 - Introduction to R]{Microeconometrics \\ Week 1 - Introduction to R}}
  - \input{header.tex}

---



```{r setup, include=FALSE}

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  #ifelse(options$size != "a", paste0("\n \\", "tiny","\n\n", x, "\n\n \\normalsize"), x)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

knitr::opts_chunk$set(echo = FALSE, dev = "png") # NOTE: switched to png instead of pdf to decrease size of the resulting pdf

library(tidyverse)
library(nycflights13)
library(kableExtra)

# read in the data
data <- read_csv("data.csv")

```



# Introduction

## Goal for today


\vfill

- The goal for today is to give you a brief introduction to R and R Markdown
\vfill

- We will be using two small datasets to get you familiar with the program
  - Class website
\vfill

- A note: if you are completely new to R, the first few weeks will be a slog
  - It will get better, I promise
\vfill

- Much of the material covered today comes from two (free!) sources:
  - [\textcolor{kdisgreen}{R for Data Science}](https://r4ds.hadley.nz/)
  - [\textcolor{kdisgreen}{R Markdown: The Definitive Guide}](https://bookdown.org/yihui/rmarkdown/)
\vfill




## What are R and RStudio?

\vfill
- R is a commonly used statistical program (and language)
  - It is free and open source, which means you can use this after graduation, without paying for it
  - R is CaSe SeNsItIvE
\vfill
- To work with R, we want to use an accompaniment called RStudio
  - RStudio is what is referred to as an integrated development environment (IDE)
  - It is not the only option (I use VS Code, for example), but it is the most common
  - It makes working with R much easier
\vfill
- Whenever you start R, you want to start RStudio
  - RStudio will start R for you
\vfill




## Some important considerations

\vfill
- One of our goals is to make **reproducible** research
  - This means that we want to be able to share our code and have others be able to replicate our results
  - To do this, we will use "scripts" that contain our code

\vfill
- A script should be self contained
  - This means that it should contain all of the code necessary to run the analysis
  - A well-written script should allow me to do everything without any additional information

\vfill
- We will also use R Markdown to create documents
  - R Markdown is a way to combine text and code
  - This allows us to create documents that are reproducible
  - We will use R Markdown to create our homework assignments
  - More on this in a bit
\vfill




## The RStudio interface

```{r rstudio1, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rstudio1.png",
                        auto_pdf = TRUE
                        )

```




## The RStudio interface

```{r rstudio2, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rstudio2.png",
                        auto_pdf = TRUE
                        )

```




## The RStudio interface

```{r rstudio3, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rstudio3.png",
                        auto_pdf = TRUE
                        )

```




## The RStudio interface

```{r rstudio4, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rstudio4.png",
                        auto_pdf = TRUE
                        )

```




## But we're missing something... what is it?
\pause

```{r rstudio5, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rstudio5.png",
                        auto_pdf = TRUE
                        )

```




## The script


```{r rstudio6, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rstudio6.png",
                        auto_pdf = TRUE
                        )

```




## Some notes
\vfill
- You can add comments to your script using a hashtag (#)
  - At the top of ALL my scripts, I have a comment that says what the script does.
  - At the top of your script, write a comment. It should say "# Week 1 - Introduction to R"
  - I put LOTS of comments in my scripts. This is good practice.
\vfill
- You can run a line of code by clicking the "Run" button
  - There are also shortcuts. On Mac it is command + enter. On windows it is control + enter. You can change these if you want.
\vfill
- You can run multiple lines of code by highlighting them and clicking the "Run" button (or the shortcut)
\vfill
- We will practice these later
\vfill



# R Basics

## Object types

- R has a few different types of objects
  - The most common are vectors, matrices, and data frames
    - A "tibble" is a type of data frame used by the tidyverse package (more below)
  - We will use data frames almost exclusively since we are working with datasets, but vectors are common, too
- You can create a vector using the c() function:
  - Note how we create a new object using the assignment operator, <-. You can also use =.

\vfill
```{r vector, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
vec <- c(1, 2, 3, 4)
vec
```



## Object types

- You can check what type of object something is by using the class() function
  - For example, if I want to check what type of object vec is, I would write class(vec)
  - Note that the output is "numeric"
  - This is because vec is a vector of numbers
- If I want to check whether it is a vector, I can write is.vector(vec)
  - Note that the output is TRUE

\vfill
```{r vector2, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
vec <- c(1, 2, 3, 4)
class(vec)
is.vector(vec)
```






## First things first: the working directory

- The working directory is the folder that R is currently working in
  - This is where R will look for files
  - This is where R will save files
  - This is where R will create files

- You can always write out an entire file path, but this is tedious
  - More importantly, it makes your code less reproducible since the path is specific to YOUR computer

- One nice thing about R is that the working directory will automatically be where you open the script from
  - Let's try this. Save your script to a folder on your computer, then open the script from that folder.





## First things first: the working directory

The working directory should be where you opened the file from. Check it like this:

```{r wd, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}

getwd()

```





## R packages

\vfill
- R is a language that is built on packages
  - Packages are collections of functions that do specific things
  - R comes with a set of "base" packages that are installed automatically
\vfill
- We are going to use one package consistently, called the "tidyverse"
  - This consists of a set of packages that are designed to work together, with data cleaning in mind
\vfill





## R packages

\vfill
The one exception to always using a script? I install packages in the CONSOLE. You can install packages like this:
\vfill
```{r tidyinstall, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny"}

install.packages("tidyverse")

```

\vfill





## Loading R packages in your script

\vfill
We need to load any R packages we want to use at the very top of the script. You should have a comment on line one, so on line two write:
\vfill
```{r tidyverse, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny"}

library("tidyverse")

```
\vfill
This will load the tidyverse package.
\vfill





## Loading data

\vfill
- Go to the class website and download the data for today.
  - Put it in your WORKING DIRECTORY (where the script is)
\vfill
- We will use the read_csv() function to load the data
  - This function is part of the tidyverse package
  - It will create a data frame
  - We need to NAME the object (data frame). As before, note the assignment operator (<-). You can actually use = though.
\vfill
```{r data, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}

library(tidyverse)

# read in the data
data <- read_csv("data.csv")

```
\vfill




## Objects in memory

The data frame should show up in the upper right hand corner of RStudio.

```{r rstudio7, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rstudio7.png",
                        auto_pdf = TRUE
                        )

```




## Objects in memory

Click on the arrow and it will show more information.

```{r rstudio8, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rstudio8.png",
                        auto_pdf = TRUE
                        )

```




## Objects in memory

\vfill
- The data frame is a matrix
  - Each row is an observation and each column is a variables
\vfill
- We can also see the names of the columns like this:
\vfill
```{r colnames, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}

colnames(data)

```
\vfill
- This is the kind of thing I might do in the console since it's not really required for the script.
\vfill





## Calling variables in R

- Some of you might be used to Stata
- One big difference between the two is that Stata generally only has one data frame in memory at a time
  - This means that you can call a variable *without referencing the data frame*
- In R, if you want to look at a variable, you have to tell R which data frame it is in
  - This is done with the $ operator
  - For example, if I want to look at the variable "age" in the data frame "data", I would write data$age
  - Let's look at summary statistics for age:
\vfill
```{r callingvar, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}

summary(data$age)

```





## Summary statistics for the entire data frame

- You can also use summary on the data frame instead of a single column
  - It helps to think of a data frame as rows and columns. For variables, you want to call specific columns.
- Look at the difference here (it cuts off because of the size of the slide):
\vfill
```{r sumstats, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}

summary(data)

```





## Calling rows/columns of a data frame (matrix)

- Think about how we refer to rows and columns in a matrix.
  - We use the row and column number, in that order.
  - For example, if I want the first row and second column of a matrix $X$, mathematically I could write $X_{1,2}$
- We do *the same thing in R*
- If I want the first row and second column of the data frame "data", I would write data[1,2]
  - Note that we use square brackets instead of parentheses
  - Note that we use a comma to separate the row and column
\vfill
```{r rowcolumn, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}

data[1,2]

```




## Calling columns of a data frame (matrix)

- We can call entire columns of a data frame by leaving the row blank
  - For example, if I want the second column of the data frame "data", I would write data[,2]
  - Note that the second column is the ability variable
\vfill
```{r column, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}

colnames(data)
data[,2]

```




## Missing variables R

- Missing variables are denoted by NA
  - This is different from Stata, which uses a period (.)
- Note that this is only how the PROGRAM stores missing variables. Sometimes the data itself has different missing values. \textcolor{red}{PAY ATTENTION!}
- For example, take a look at the first ten rows of the data frame (also note how I call the first ten rows and leave out the first column!):
\vfill
```{r missing, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
data[1:10,-1]
```




## Variable types

- R also has a few different types of variables
  - The most common are numeric, character, and logical
- Look at the previous code again:
\vfill
```{r missing2, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
data[1:10,-1]
```




## Variable types

- dbl is short for double, which is a numeric variable (the "type" of numeric variable is about how much memory is needed to store it)
- chr is short for character, which is a string of characters (text)
  - Surprisingly, in our previous example, educyears was a character string even though it seemed to be a number
  - Let's look at the possible values of educyears using the unique() function, which outputs a vector:
\vfill
```{r uniqueeducyears, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
unique(data$educyears)
```




## Variable types

- Interesting! It seems that there is a "Not Mentioned" value.
  - What if we want to replace those with missing, instead?
- Let's talk through the following code
  - First note how it refers to a specific column and then a specific row
  - Also note how it uses two equal signs (==) to check whether the value is "Not Mentioned"
    - This is similar to Stata!
\vfill
```{r replace, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
# replace "Not Mentioned" with NA
data$educyears[data$educyears == "Not Mentioned"] <- NA  
# check that it worked by looking at the unique values
unique(data$educyears)              
# turn into numeric
data$educyears <- as.numeric(data$educyears)
class(data$educyears)
```






## Pipes

- One of the most useful things in R is the pipe operator (%>%)
  - This is part of the tidyverse package
  - It allows you to chain commands together
  - It makes your code much easier to read
  - It makes your code much easier to write
  - It makes your code much easier to debug
  - It makes your code much easier to share
  - It makes your code much easier to reproduce
\vfill
- It's easy to use but it will take some time for you to get used to the names of the functions we can use with it
  - This also goes for other tasks in R, not just with the pipe operator






## Pipes example
Here is an example of how we can use pipes with the mutate() function in tidyverse

- We are also going to use ifelse() to make this work
\vfill
```{r replacemutate, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
data <- data %>%
          mutate(educyears = ifelse(educyears == "Not Mentioned", NA, educyears), # if educyears=="Not Mentioned", replace
                educyears = as.numeric(educyears))    # replace educyears as numeric (instead of character)
summary(data$educyears)
```
\vfill\pause
Note that we could wrap as.numeric() around the ifelse() command to do it on one line!
\vfill
```{r replacemutate2, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears))) # wrapped into one line
summary(data$educyears)
```







## Missings and functions in R
In Stata, by default, functions ignore missing values

- R does not do this by default. Look at this:
\vfill
```{r missingsNA, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears))) # wrapped into one line
mean(data$educyears)
```
\vfill
If there are any missing values, the function will evalute to missing!

- But we can also do this:
\vfill
```{r missingsNA2, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears))) # wrapped into one line
mean(data$educyears, na.rm = TRUE) # BE CAREFUL WITH THIS! Make sure it is indeed what you want to do.
```






## Functions and storing values
The mean() function in the previous slide outputs a single value
- That means we could store that value as an object:
\vfill
```{r storingvalues, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears))) # wrapped into one line
meaneduc <- mean(data$educyears, na.rm = TRUE)
sdeduc <- sd(data$educyears, na.rm = TRUE)
meaneduc
sdeduc
```
\vfill
How is this helpful? We can use these values later in our script!






## Functions and mutate()
We can combine the mean() and sd() functions within mutate to create a new, standardized variable:
\vfill
```{r mutate_std, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"} 
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears)), # wrapped into one line
                 educyears_std = (educyears - mean(educyears))/sd(educyears))
summary(data$educyears_std)
```
\vfill
Oh no! what happened?






## Functions and mutate()
We can combine the mean() and sd() functions within mutate to create a new, standardized variable:
\vfill
```{r mutate_std2, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"} 
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears)), # wrapped into one line
                 educyears_std = (educyears - mean(educyears, na.rm = T))/sd(educyears, na.rm = T))
summary(data$educyears_std)
```
\vfill
Note that we can shorten TRUE to T (or FALSE to F).






## Visualizations with ggplot2
- ggplot2 is a flexible way to create visualizations in R
- The basic idea is that you create a plot object and then add layers to it
- Let's create a histogram of educyears






## Visualizations with ggplot2
```{r ggplot, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "45%", fig.align = "center"}
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears)))
# we call ggplot() and NOT ggplot2()
ggplot() +   # note how we use + here, NOT the pipe operator
  geom_histogram(data = data, aes(x = educyears)) # the histogram with geom_histogram
# data = data tells R to use the data frame "data", and the aes() is the aesthetic
# only an x value here since a histogram uses just a SINGLE value
```






## Visualizations with ggplot2
```{r ggplot2, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "45%", fig.align = "center"}
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears)))
# we can save the plot as an object
g1 <- ggplot() +
        geom_histogram(data = data, aes(x = educyears))
g1
```






## Visualizations with ggplot2
```{r ggplot3, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "45%", fig.align = "center"}
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears)))
# lots of ways to change the plot
g1 <- ggplot() +
        geom_histogram(data = data, aes(x = educyears)) +
        labs(title = "Histogram of educyears",
             x = "Years of education",
             y = "Count")
g1
```






## One more example
```{r ggplot4, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "45%", fig.align = "center"}
data <- data %>%
          mutate(educyears = as.numeric(ifelse(educyears == "Not Mentioned", NA, educyears)))
g1 <- ggplot() +
        geom_histogram(data = data, aes(x = educyears)) +
        labs(title = "Histogram of educyears",
             x = "Years of education",
             y = "Count") +
        theme_minimal()
g1
```






## Let's try this with a NEW dataset
First install a new package that has a dataset we will use (you can do this in the console):

```{r install, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny", out.width = "45%", fig.align = "center"}
install.packages("nycflights13")
```
\vfill
Now let's see:

```{r flights, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "45%", fig.align = "center"}
library(nycflights13)
glimpse(flights)
```






## Let's look at some new tidyverse functions
\vfill
Let's get the average departure delay by NYC airport:
\vfill
```{r flights2, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
flights %>% 
    group_by(origin) %>% # this groups ROWS based on their origin value
    summarize(avg_dep_delay = mean(dep_delay, na.rm = T)) # this summarizes the data, creating means absed on the grouping!
```
\vfill
Note that this does not create a single value. Instead it creates a tibble (a data frame) summarizing the data by our grouping variable.
\vfill






## Let's look at some new tidyverse functions
\vfill
What if we want to save that tibble instead?
\vfill
```{r flights3, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny"}
summat <- flights %>% 
            group_by(origin) %>% # this groups ROWS based on their origin value
            summarize(avg_dep_delay = mean(dep_delay, na.rm = T)) # this summarizes the data, creating means based on groups!
summat # print the 3x2 matrix in the console
```
\vfill
I could then output this to a table if I wanted to (using Markdown, more on this later):
```{r flights4, echo = FALSE, message = FALSE, warning = FALSE, size = "tiny"}
summat <- flights %>% 
            group_by(origin) %>% # this groups ROWS based on their origin value
            summarize(avg_dep_delay = mean(dep_delay, na.rm = T)) # this summarizes the data, creating means based on groups!
kable(summat,
      align = "c", linesep = "") %>%
      kable_classic_2()
```






## Let's look at a new plot
How does departure delay vary by time of day?
\vfill
```{r flights5, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
ggplot() + 
  geom_smooth(data = flights, aes(x = sched_dep_time, y = dep_delay))
```






## Let's look at a new plot
We can color code by origin, too!
\vfill
```{r flights6, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
ggplot() + 
  geom_smooth(data = flights, aes(x = sched_dep_time, y = dep_delay, color = origin))
```






## Make it prettier
```{r flights7, echo = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
ggplot() + 
  geom_smooth(data = flights, aes(x = sched_dep_time, y = dep_delay, color = origin), se = FALSE) +
  labs(x = "Scheduled departure time",
       y = "Departure delay (minutes)") +
  theme_minimal() + guides(color = guide_legend(title = "Departure airport"))
```





# R Markdown

## What is R Markdown?

- R Markdown is a way to combine text and code
  - This allows us to create documents that are reproducible
  - We will use R Markdown to create our homework assignments

- These slides were all created in R Markdown

- My papers are written in R Markdown (well, some of them are, anyway)
  - [\textcolor{kdisgreen}{\textbf{Here}}](https://joshmerfeld.github.io/assets/papers/pollution2023.pdf) is an example

- Yihui Xie, J. J. Allaire, and Garrett Grolemund have an awesome -- free! -- resource on R Markdown, [\textcolor{kdisgreen}{\textbf{R Markdown: The Definitive Guide}}](https://bookdown.org/yihui/rmarkdown/)





## Installing R Markdown
\vfill
You'll need to install R Markdown. You can do this in the console:
\vfill
```{r installrmarkdown, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny"}
install.packages("rmarkdown")
```
\vfill





## Creating an R Markdown document in RStudio

```{r rm1, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rm1.png",
                        auto_pdf = TRUE
                        )

```





## Creating an R Markdown document in RStudio

```{r rm2, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rm2.png",
                        auto_pdf = TRUE
                        )

```





## Creating an R Markdown document in RStudio

```{r rm3, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rm3.png",
                        auto_pdf = TRUE
                        )

```





## Go ahead and save this document

- Go ahead and save this document in your working directory.
  - One think about Markdown files is that it will ALWAYS set the working directory to where the file is saved whenever you "knit" the document.

- What is "knitting"?
  - Knitting is the process of turning your R Markdown document into a pdf, html, or word document.
  - We will just focus on pdfs for now.





## Knit it!

```{r rm4, echo = FALSE, message = FALSE, warning = FALSE, out.width = "90%", fig.align = "center"}

knitr::include_graphics(
                        path = "assets/rm4.png",
                        auto_pdf = TRUE
                        )

```





## Check out the document you just created

- Go to your working directory and open the pdf to see what it looks like.
  - It will always create the pdf in the same folder as the .Rmd file.





## YAML header

\vfill
- At the very top of the document is some information about the document
  - This is called the YAML header
  - It tells R Markdown what kind of document to create
  - It also allows you to set some options
  - DO NOT DELETE THE --- AT THE TOP AND BOTTOM OF THE YAML HEADER!
\vfill  
- You can change the title and date as you please
  - For today's date, you can use Sys.Date() within R inline code (more in a second):
\vfill
```{r date, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
date: "`r Sys.Date()`"
```
\vfill





## The setup chunk

- Just below the YAML header you'll see a "code chunk" called "setup" (r setup, include = FALSE)
  
- Note how it has $```$ and $```$ at the top and bottom. This differentiates the "code chunk" from the rest of the document.
  - Whenever you want to add a code chunk, you *must* have the $```$ at the top and bottom of it, at the beginning of the line.

- Use the setup code chunk to load any packages or data that you want to use in the rest of the document.
  - Later code chunks are "local": they will be able to access things from the setup chunk *but not from other code chunks*.





## The setup chunk
This is an example of what the setup chunk looks like.

````markdown

`r ''````{r setup, include=FALSE}
# universal chunk options.
# echo = TRUE will show the code in the document.
# echo = FALSE will not.
knitr::opts_chunk$set(echo = TRUE)  

# load any packages you want to use throughout the document.
library(tidyverse) 

# load any data you want to use throughout the document.
data <- read_csv("data.csv") 
```
````





## Code chunks
Here is an example of a regular code chunk.

````markdown

`r ''````{r chunkexample, include = TRUE}
# note that I named the chunk.
# all chunks must have a UNIQUE name.
# you will get an error if they don't

# I already loaded by data above
ggplot(flights) + 
  geom_histogram(aes(x = air_time), binwidth = 10)
```
````





## Code chunks
Here is the output of that chunk:

```{r chunkexample, include = TRUE}
# note that I named the chunk.
# all chunks must have a UNIQUE name.
# you will get an error if they don't

# I already loaded by data above
ggplot(flights) + 
  geom_histogram(aes(x = air_time), binwidth = 10)
```





## Code chunks
Oh no! It looks bad! Changes:

```{r chunkexample2, include = TRUE, warning = FALSE, out.width = "55%", fig.align = "center"}
# note that I named the chunk.
# all chunks must have a UNIQUE name.
# you will get an error if they don't

# I already loaded by data above
ggplot(flights) + 
  geom_histogram(aes(x = air_time), binwidth = 10)
```





## Code chunks
How did I do that?

\small
````markdown
`r ''````{r chunkexample, include = TRUE, warning = FALSE, out.width = "55%", 
          fig.align = "center"}
# note that I named the chunk.
# all chunks must have a UNIQUE name.
# you will get an error if they don't

# I already loaded by data above
ggplot(flights) + 
  geom_histogram(aes(x = air_time), binwidth = 10)
```
````

- NOTE: The start of the chunk must be on ONE line. It is wrapped here just for presentation.




## Code chunks options

- There are lots of code chunk options.

- You can find the code chunk options [\textcolor{kdisgreen}{here}](https://rpubs.com/Lingling912/870659) (https://rpubs.com/Lingling912/870659)

- This will get easier to use as you get more and more practice.





## Starting new sections/subsections

````markdown
# This will create a new section

## This will create a new sub-section

### This will create a new sub-sub-section
Don't do this.
````





## You can add R inline code

\vfill
- You can add R inline code using the $`r'$ operator.
  - For example, if I want to add the date, I can write `r Sys.Date()`
  - There are `r ncol(flights)` columns in the flights data.
  - There are `r nrow(flights)` columns in the flights data.

\vfill
````markdown
- You can add R inline code using the $`r'$ operator.
  - For example, if I want to add the date, I can write `r Sys.Date()`
  - There are `r ncol(flights)` columns in the flights data.
  - There are `r nrow(flights)` columns in the flights data.
````
\vfill




## Enumerated lists/bullets
\vfill
- I like lists.
  - With indentations.
1. Really, I do.
   1. Indent! It's just a tab.
\vfill
````markdown
- I like lists.
  - With indentations.
1. Really, I do.
   1. Indent! It's just a tab.
````
\vfill




## latex

- R Markdown uses latex to create pdfs. This allows you to do some cool things.

- For example, it is easy to add equations with latex, using \$:

$y = x + \varepsilon$
\vfill
````markdown
- For example, it is easy to add equations with latex, using \$:

$y = x + \varepsilon$
````




## latex

- I can center it, too:

$$y = x + \varepsilon$$ 
\vfill
````markdown
- I can center it, too:

$$y = x + \varepsilon$$ 
````




## latex

\begin{gather}\label{eq1} y = x + \varepsilon \end{gather}
- In Equation \autoref{eq1}

\vfill
````markdown
\begin{gather}\label{eq1} y = x + \varepsilon \end{gather}
- In Equation \autoref{eq1}
````

\vfill
- You might think, so what? Well what's cool is that if we add equations before it, the number will automatically update!





## latex

- latex is particularly helpful for rendering math
  
- You can find a handy reference guide [\textcolor{kdisgreen}{here}](https://icl.utk.edu/~mgates3/docs/latex.pdf) (https://icl.utk.edu/~mgates3/docs/latex.pdf)





## Creating tables

- There are lots of ways to create tables in R Markdown.
  - I will show you how using the kable() function in the knitr package.
  - You do not need to download this package, it is already installed with R Markdown.
    - There is extra functionality in the kableExtra package. You need to download this and laod it if you want to use it.





## Creating tables

```{r table, include = TRUE, eval = FALSE, warning = FALSE, echo = TRUE, size = "tiny"}
summat <- flights %>% 
            # this groups ROWS based on their origin value
            group_by(origin) %>% 
            # create means by group
            summarize(avg_dep_delay = mean(dep_delay, na.rm = T),
                      avg_arr_delay = mean(arr_delay, na.rm = T),
                      avg_air_time = mean(air_time, na.rm = T),
                      flights = n())
# output 
kable(summat,
      align = "cccc")
```





## Creating tables

```{r table2, include = TRUE, eval = TRUE, warning = FALSE, echo = FALSE, size = "tiny"}
summat <- flights %>% 
            # this groups ROWS based on their origin value
            group_by(origin) %>% 
            # create means by group
            summarize(avg_dep_delay = mean(dep_delay, na.rm = T),
                      avg_arr_delay = mean(arr_delay, na.rm = T),
                      avg_air_time = mean(air_time, na.rm = T),
                      flights = n())
# output 
kable(summat,
      align = "ccc")
```





## I don't like that at all! Let's make it pretty.

```{r table3, include = TRUE, eval = FALSE, warning = FALSE, echo = TRUE, size = "tiny"}
summat <- flights %>% 
            # this groups ROWS based on their origin value
            group_by(origin) %>% 
            # create means by group, ROUNDING to two decimal places
            summarize(avg_dep_delay = round(mean(dep_delay, na.rm = T), 2),
                      avg_arr_delay = round(mean(arr_delay, na.rm = T), 2),
                      avg_air_time = round(mean(air_time, na.rm = T), 2),
                      flights = n())
# rename columns
colnames(summat) <- c("Origin", "Departure Delay", "Arrival Delay", "Flight Time", "Flights")
# output 
kable(summat, caption = "Averages by origin (minutes)",
      align = "ccc", linesep = "", 
      booktabs = TRUE) %>% # this is from kablextra. You don't have to use it, but I like it.
      kable_classic_2() # this is also from kablextra
```





## I don't like that at all! Let's make it pretty.

```{r table4, include = TRUE, eval = TRUE, warning = FALSE, echo = FALSE, size = "tiny"}
summat <- flights %>% 
            # this groups ROWS based on their origin value
            group_by(origin) %>% 
            # create means by group
            summarize(avg_dep_delay = round(mean(dep_delay, na.rm = T), 2),
                      avg_arr_delay = round(mean(arr_delay, na.rm = T), 2),
                      avg_air_time = round(mean(air_time, na.rm = T), 2),
                      flights = n())
# rename columns
colnames(summat) <- c("Origin", "Departure Delay", "Arrival Delay", "Flight Time", "Flights")
# output 
kable(summat, caption = "Averages by origin (minutes)",
      align = "ccc", linesep = "", 
      booktabs = TRUE) %>% # this is from kablextra. You don't have to use it, but I like it.
      kable_classic_2() # this is also from kablextra
```





## One more change!

```{r table5, include = TRUE, eval = FALSE, warning = FALSE, echo = TRUE, size = "tiny"}
summat <- flights %>% 
            # this groups ROWS based on their origin value
            group_by(origin) %>% 
            # create means by group, ROUNDING to two decimal places
            summarize(avg_dep_delay = round(mean(dep_delay, na.rm = T), 2),
                      avg_arr_delay = round(mean(arr_delay, na.rm = T), 2),
                      avg_air_time = round(mean(air_time, na.rm = T), 2),
                      flights = n())
summat$flights <- format(summat$flights, big.mark = ",", scientific = FALSE)
# rename columns
colnames(summat) <- c("Origin", "Departure Delay", "Arrival Delay", "Flight Time", "Flights")
summat <- t(summat)
# output 
kable(summat, caption = "Averages by origin (minutes)",
      align = "ccc", linesep = "", 
      booktabs = TRUE) %>% # this is from kablextra. You don't have to use it, but I like it.
      row_spec(c(1, 4), hline_after = TRUE) %>% # this is also from kablextra
      kable_classic_2() # this is also from kablextra
```





## One more change!

```{r table6, include = TRUE, eval = TRUE, warning = FALSE, echo = FALSE, size = "tiny"}
summat <- flights %>% 
            # this groups ROWS based on their origin value
            group_by(origin) %>% 
            # create means by group
            summarize(avg_dep_delay = round(mean(dep_delay, na.rm = T), 2),
                      avg_arr_delay = round(mean(arr_delay, na.rm = T), 2),
                      avg_air_time = round(mean(air_time, na.rm = T), 2),
                      flights = n())
summat$flights <- format(summat$flights, big.mark = ",", scientific = FALSE)
# rename columns
colnames(summat) <- c("Origin", "Departure Delay", "Arrival Delay", "Flight Time", "Flights")
summat <- t(summat)
# output 
kable(summat, caption = "Averages by origin (minutes)",
      align = "ccc", linesep = "", 
      booktabs = TRUE) %>% # this is from kablextra. You don't have to use it, but I like it.
      row_spec(c(1, 4), hline_after = TRUE) %>% # this is also from kablextra
      kable_classic_2() # this is also from kablextra
```





## Enough for now

- That's enough on tables for now

- As you can see, there are lots of ways to customize tables

- Where this becomes really powerful is when you combine it with R code to create tables dynamically
  - I will teach you to use a package called fixest that helps automate some of this
  - If you change your specification, your tables will update AUTOMATICALLY!
  - Ever tried to manually change a table in Word? Never again.





## Some tips

- When I write a paper in Markdown, I generally do not do all of my analysis in the Markdown document
  
- Instead, I do the analysis in another script and then save the resulting tables
  
- I then load these tables in the setup chunk of my Markdown document and use them in the document
  - For figures, it depends. For a simple summary figure, I might load the data in the Markdown document and create the figure there.





## First assignment

- Assignment for next week (due one week from today):
  - Create a simple markdown document
  - You can find the assignment [\textcolor{kdisgreen}{here}](https://github.com/JoshMerfeld/applied-microeconometrics/blob/main/assignments/assignment1/assignment1.pdf) (on the [\textcolor{kdisgreen}{course GitHub page}](https://github.com/JoshMerfeld/applied-microeconometrics))

- Next week, you will turn in on e-KDIS:
  - R script (if there is one)
  - R Markdown script
  - pdf of the R Markdown document





