---
title:  |
    | Microeconometrics (Causal Inference)
    | Week 4 - Introduction to causality
author:
  |
    | Joshua D. Merfeld
    | KDI School of Public Policy and Management
date: "`r Sys.Date()`"

# Output type and options
output: 
  beamer_presentation:
    theme: Montpellier
classoption: "aspectratio=169"

# This includes latex arguments
header-includes:
  - \AtBeginDocument{\title[Week 4 - Introduction to causality]{Microeconometrics (Causal Inference) \\ Week 4 - Introduction to causality}}
  - \AtBeginDocument{\author[Josh Merfeld - KDI School]{Joshua D. Merfeld \\ KDI School of Public Policy and Management}}
  - \input{header.tex}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = "png") # NOTE: switched to png instead of pdf to decrease size of the resulting pdf

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  #ifelse(options$size != "a", paste0("\n \\", "tiny","\n\n", x, "\n\n \\normalsize"), x)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})




library(tidyverse)
library(kableExtra)
library(fixest)
library(ggpubr)
library(RColorBrewer)
library(haven)
library(mfx)
library(nnet)
library(survival)
library(survminer)
library(medicaldata)
library(MatchIt)
  
licresults <- feols(pacu30min_throatPain ~ treat, data = licorice_gargle)

df <- read_csv("matchingdata.csv")


```



# Introduction

## What are we doing today?
\vfill

- This week we will discuss causation
\vfill

- For example, what prevents us from saying that something is a causal effect?
\vfill

- Then:
  - How do RCTs help?
  - Propensity scores
\vfill




## What are we doing today?
or...\pause
\vfill
\center \textit{Correlation is not (necessarily) causation}
\vfill




## Let me start with an example

\vfill
- I used to live in Atlanta
  - There was a large hospital in downtown: Grady
\vfill
- Grady was a trauma center
  - It was the only one in the immediate area
\vfill




## Let me start with an example
\vfill
- Something I once heard: Mortality rates at Grady are so much higher than at other hospitals. Is it a bad hospital?
\vfill
- What do you think? Is Grady necessarily a worse hospital?
\vfill







# Potential outcomes

## Potential outcomes framework
\vfill
- Let's introduce one of the most common ways to think about causality (in economics): the potential outcomes framework
  - This framework is also known as the Rubin Causal Model
\vfill
- The potential outcomes framework is a way to think about causality
  - It is not the only way
  - It is not necessarily the best way
  - But it's what we're going to use
\vfill






## Potential outcomes framework
\vfill
- Suppose someone is in a car accident
  - They could be taken to Grady or another hospital
\vfill
- Grady is the treatment:
  - $D = 1$ if they go to Grady
  - $D = 0$ if they go to another hospital
\vfill
- We can think of the outcome of interest as the person's health
  - We can think of this as a \textit{potential outcome}
  - We can think of the person's health if they go to Grady as $Y_{1}$
  - We can think of the person's health if they go to another hospital as $Y_{0}$
\vfill






## Potential outcomes framework
\vfill
- Let's write this all out, imagining different possible people, indexed by $i$:
\vfill
\begin{gather} \text{Potential outcomes} = \Biggl\{ \begin{array}{ll} Y_{1i} & \text{if } D_i=1 \\ Y_{0i} & \text{if } D_i=0 \end{array} \end{gather}
\vfill

- $Y_{1i}$ is the outcome for person $i$ if they go to Grady
- $Y_{0i}$ is the outcome for *the same person at the same time* if they go to another hospital
\vfill





## What's the problem?
\vfill
\begin{gather} \text{Potential outcomes} = \Biggl\{ \begin{array}{ll} Y_{1i} & \text{if } D_i=1 \\ Y_{0i} & \text{if } D_i=0 \end{array} \end{gather}
\vfill

- $Y_{1i}$ is the outcome for person $i$ if they go to Grady
- $Y_{0i}$ is the outcome for *the same person at the same time* if they go to another hospital
\vfill

- What's the problem we have?
\vfill





## What's the problem?
\vfill
\center
\textit{We never observe the same person at the same time in two different states of nature \\ (i.e. going to Grady AND going to another hospital at the same time)}
\vfill





## Let's just compare those who go to Grady with those who don't?
- Actual causal effect:
$$\mathbb{E}(Y_{1i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=0)$$
\vfill
- Comparing across individuals:
$$\mathbb{E}(Y_{i}|D_i=1) - \mathbb{E}(Y_{i}|D_i=0)$$
\vfill
- Note that this is *not* the same thing. We are comparing different people.





## Let's just compare those who go to Grady with those who don't?
- We can break down this comparison into two separate terms:
\begin{align}\mathbb{E}(Y_{i}|D_i=1) - \mathbb{E}(Y_{i}|D_i=0) =& \nonumber \\
              &\left[\mathbb{E}(Y_{1i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=1)\right]  \\
               +&\left[\mathbb{E}(Y_{0i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=0)\right] \end{align}





## Let's just compare those who go to Grady with those who don't?
\begin{align}\mathbb{E}(Y_{i}|D_i=1) - \mathbb{E}(Y_{i}|D_i=0) =& \nonumber \\
              &\left[\mathbb{E}(Y_{1i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=1)\right] \tag{3} \\
               +&\left[\mathbb{E}(Y_{0i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=0)\right] \tag{4} \end{align}

- Line 3 is the causal effect we're interested in: 
$$\mathbb{E}(Y_{1i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=1) = \mathbb{E}(Y_{1i}-Y_{0i}|D_i=1)$$

  - This is the *treatment effect of going to Grady on those who went to Grady*
  - Their actual outcome minus their counterfactual outcome





## Let's just compare those who go to Grady with those who don't?
\begin{align}\mathbb{E}(Y_{i}|D_i=1) - \mathbb{E}(Y_{i}|D_i=0) =& \nonumber \\
              &\left[\mathbb{E}(Y_{1i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=1)\right] \tag{3} \\
               +&\left[\mathbb{E}(Y_{0i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=0)\right] \tag{4} \end{align}

- Line 4 is the real problem. This is *selection bias*.
  - This is the difference between the potential outcomes for those who went to Grady and those who didn't.
  - The question: What might be the *difference* between people who go to Grady and people who go to another hospital?
  - If you ever hear an economist say "selection", this is what they're talking about: the systematic differences between the two groups.





## Who goes to Grady?
\center
\includegraphics[width = 0.6\columnwidth]{assets/grady1.png}





## Who goes to Grady?
\center
\includegraphics[width = 0.6\columnwidth]{assets/grady2.png}





## Who goes to Grady?
\center
\includegraphics[width = 0.6\columnwidth]{assets/grady3.png}





## Who goes to Grady?
\vfill  
- So if people with much worse injuries are going to Grady, then we might expect that the people who go to Grady have worse outcomes...
\vfill  
- Even if Grady is better at treating bad injuries!
\vfill  







# RCTs

## So what to do?
\vfill
- The rest of this course is about ways to address this problem
\vfill
- Let's start with the "gold standard": randomized controlled trials (RCTs)
\vfill





## RCTs
\vfill
- So how do RCTs help?
\vfill 
- RCTs can't help with the fact that we can never observe the same unit in both treatment states at the same time
\vfill
- Instead, RCTs rely on groups
\vfill





## RCTs
\vfill
- Let's go back to our diabetes/retinopathy example from last week
\vfill
- Why wouldn't we want to simply compare individuals who received a new laser eye treatment for retinopathy to people who don't?
\vfill





## Random assignment and selection
\vfill
- The key is random assignment
\vfill
- Random assignment means that, *on average*, those who receive treatment and those who do not are the same!
\vfill
- Our selection bias equation is:
\begin{gather} \mathbb{E}(Y_{0i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=0) \end{gather}
\vfill

- Guess what this equals when we have random assignment?
\vfill





## Random assignment and selection
\newcommand{\indep}{\perp \!\!\! \perp}
- Mathematically, random assignment means that potential outcomes are *independent* of treatment assignment:
\begin{gather} \{Y_{1i}, Y_{0i}\} \indep D_i \end{gather}

- This allows us to do the following:
\begin{align} \text{selection} &= \mathbb{E}(Y_{0i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=0) \\
                               &= \mathbb{E}(Y_{0i}|D_i=1) - \mathbb{E}(Y_{0i}|D_i=1)  \end{align}

- The last line is zero! In expectation, there is no selection bias when treatment is randomized.






## RCTs

- We're going to use a new dataset to illustrate the analysis of RCTs.
  - We need to download the package from GitHub!
\vfill
```{r rct1, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
install.packages("remotes") # install package remotes
remotes::install_github("higgi13425/medicaldata") # download datasets from github repo
# repo here: https://higgi13425.github.io/medicaldata/
library(medicaldata) # load datasets
```

\vfill






## RCTs

\includegraphics[width = \columnwidth]{assets/rct1.png}
\vfill
```{r rct2, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
colnames(licorice_gargle)
```








## Analyzing RCTs
\vfill
- Analyzing RCTs with simple randomization is easy!
\vfill
- We can simply compare the average outcome for those who received treatment to the average outcome for those who did not:
$$ y_{i} = \beta_{0} + \beta_{1}D_{i} + \epsilon_{i}, $$
where $y_{i}$ is the outcome of interest and $D_{i}$ is a dummy variable for treatment.
\vfill






## Effect of licorice gargle on sore throat

```{r rct3, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
licresults <- feols(pacu30min_throatPain ~ treat, data = licorice_gargle)
summary(licresults)
etable(licresults, licresults, 
      vcov = list("iid", "HC1"),
      digits = 3,
      headers = c("sore throat", "sore throat"),
      depvar = FALSE)
```






## Make it a pretty table

```{r rct4, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
etable(licresults, licresults, 
      vcov = list("iid", "HC1"),
      digits = 3,
      se.below = TRUE, # add SEs BELOW coefficients (the norm)
      depvar = FALSE)
```






## Even nicer

```{r rct5, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
table <- etable(licresults, licresults, 
          vcov = list("iid", "HC1"),
          digits = 3, digits.stats = 3,
          se.below = TRUE, # add SEs BELOW coefficients (the norm)
          depvar = FALSE) # don't list dep var at top
table <- table[-c(5:6, 9:11), ]
colnames(table) <- c("", "(1)", "(2)")
kable(table,
          caption = "Effect of licorice gargle on sore throat",
          booktabs = TRUE, align = "lcc", linesep = "", escape = FALSE, row.names = FALSE) %>%
          footnote("* p<0.1 ** p<0.05 *** p<0.01", general_title = "", footnote_as_chunk = TRUE) %>%
          row_spec(4, hline_after = TRUE) %>%
          kable_classic_2()
          
```










## Controlling for variables
\vfill
- We often have baseline values of the outcome of interest
  - For example, in an intervention about income, we have a baseline measure of income
\vfill
- It is quite common to use baseline values as controls in regressions
  - Why? It helps improve power!
\vfill
- ANCOVA, but it's really just simple OLS regression
\vfill
- Important: never control for things that can change due to treatment!
\vfill










## Some complications in analyzing RCTs
\vfill
- What if we have multiple treatment groups?
\vfill
- What if randomization is at an aggregate level?
\vfill
- What if we have multiple outcomes?
\vfill
- What if randomization depends on other variables?
\vfill






## Multiple treatment groups
\vfill
- Suppose we have three groups:
  - $D_{i} = 0$ if control
  - $D_{i} = 1$ if treatment 1
  - $D_{i} = 2$ if treatment 2
\vfill
- This is pretty common
  - Sometimes there are multiple treatments
  - Sometimes treatments are layered on top of one another
\vfill






## Fischer et al. (2023)
\center
	\includegraphics[width=0.8\columnwidth]{assets/fischer1}






## Note that they randomize *villages*, not households
\center
  \includegraphics[width=0.8\columnwidth]{assets/fischer2}






## Analyzing multiple treatment groups
\vfill
- We can still use the same equation:
$$ y_{i} = \beta_{0} + \sum_{k=1}^K\beta_{k}D_{k} + \epsilon_{i}, $$
where $y_{i}$ is the outcome of interest and $k$ indexes the different treatment groups.
\vfill
- In the simplest case, we can simply compare the average outcome for each treatment group to the average outcome for the control group.
\vfill
- We can also compare the average outcome for each treatment group to the average outcome for each other treatment group.
  - Note that we use an F-test for this: we are testing whether some of the $\beta_{k}$ are equal to *one another*, which is a coefficient restriction.
\vfill






## Aggregate randomization
\vfill
- We've talked about individual randomization
  - But in this example, randomization is at the village level
\vfill
- So how do we take this into account?
\vfill\pause
- Cluster standard errors at the level of randomization!
  - In this case, cluster standard errors at the village level
  - If you randomize schools, you would cluster at the school level
  - Etc.
\vfill






## Aggregate randomization: clustered standard errors
\center
\includegraphics[width = 0.8\columnwidth]{assets/fischer3}






## Note the use of the F-test here
\center
\includegraphics[width = 0.8\columnwidth]{assets/fischer4}







## Complication two: multiple outcomes
\vfill
- Suppose we have many outcomes
  - For argument's sake, let's say we have 10 outcomes
\vfill
- A single p-value can be misleading
  - If we have 10 outcomes and two treatments, we expect one of them to be significant at the 5\% level by chance alone!
  - The more outcomes you have, the more likely you are to find significant effects, *even if there are none*!
\vfill







## Relevant xkcd (882)
\center
\includegraphics[width = 0.8\columnwidth]{assets/xkcd1}







## Relevant xkcd (882)
\center
\includegraphics[width = 0.8\columnwidth]{assets/xkcd2}







## Relevant xkcd (882)
\center
\includegraphics[width = 0.8\columnwidth]{assets/xkcd3}







## Relevant xkcd (882)
\center
\includegraphics[width = 0.8\columnwidth]{assets/xkcd4}







## Relevant xkcd (882)
\center
\includegraphics[width = 0.8\columnwidth]{assets/xkcd5}







## Relevant xkcd (882)
\center
\includegraphics[width = 0.8\columnwidth]{assets/xkcd6}










## The problem, mathematically
\vfill
- Suppose you have one outcome
  - Suppose the null hypothesis is true
  - What is the probability of finding a significant effect at the 5\% level? What is the probability of not finding a significant effect?\pause
  - 5\% and 95\%, respectively
\vfill
- Suppose you have 20 outcomes
  - Suppose all null hypotheses are true
  - What is the probability of finding NO significant effects, assuming independence across outcomes?\pause
  - $0.95^{20} = 0.358$
\vfill










## Correcting for multiple outcomes
\vfill
- Bonferroni correction
  - This is the simplest correction
  - If you have 10 outcomes, you multiply your p-value by 10
  - If you have 20 outcomes, you multiply your p-value by 20 (cap p-values at 1.000)
\vfill
- This is very rare nowadays in applied microeconometrics
  - It's too conservative
\vfill










## Correcting for multiple outcomes
\vfill
- Sharpened q-values (Benjamini, Krieger, and Yekutieli, 2006; Anderson, 2008)
  - R package ``qvalue``
    - Just requires a vector of p-values
  - An advantage is their comparison to p-values
  - A disadvantage is that it does not take into account correlations across outcomes
\vfill 
- This is related to the False Discovery Rate (FDR)
  - Read more here: https://blogs.worldbank.org/impactevaluations/overview-multiple-hypothesis-testing-commands-stata
\vfill










## From Anderson (2008)
\center
\includegraphics[width = 0.8\columnwidth]{assets/q1}










## List et al. (2016)
\vfill
- List et al. (2016) is a good example of how to deal with multiple outcomes
  - Their procedure allows for p-values to be correlated
    - This is about familywise error rates (FWER)
  - Much more conservative as you add outcomes, because it is about avoiding *any* false positives (type one errors)
\vfill
- Unfortunately, I haven't found an R package for this yet, and we aren't going to do this by hand
  - In Stata: ``ssc install mhtexp``
\vfill
- For this class, you can use q-values
\vfill










## Stratification
\vfill
- Final complication: what if randomization depends on other variables?
\vfill
- Stratification
  - People in different "strata" have differential probability of being treated
  - Example:
    - Probability of male-owned firm being selected is 0.5
    - Probability of female-owned firm being selected is 0.75
    - In this case, we can say that selection is stratified by gender
\vfill










## Stratification
\vfill
- How do we deal with stratification?
  - By including dummy variables for the strata!
$$ y_{i} = \beta_{0} + \beta_{1}D_{i} + \sum_{s=2}^SI(G_s) + \epsilon_{i}, $$
where $s$ is the stratum number, $I(G_s)$ is an indicator for being in stratum $s$, and $S$ is the number of strata. (Note that we don't include a dummy for the first stratum.)
\vfill







# Propensity scores

## Propensity scores
\vfill
- Let's talk about propensity scores!
\vfill
- Propensity scores are a way to deal with selection bias, provided some assumptions are met
\vfill
- This section draws heavily from Elizabeth Stuart's slides: \textcolor{kdisgreen}{http://www.preventionresearch.org/wp-content/uploads/2011/07/SPR-Propensity-pc-workshop-slides.pdf}
  - We will use one of her packages: MatchIt (`install.packages("MatchIt")`), note the capitalization!
  - https://kosukeimai.github.io/MatchIt/index.html
\vfill






## Let's first talk about matching
\vfill
- The idea behind matching is to approach the experimental idea:
  - The treatment group and the control group are approximately the same
\vfill
- The catch: we can only do this for *observables*
  - While RCTs also implicitly "match" on unobservables, we can't do that here
\vfill 






## Consider the following example
\vfill
- We are interested in the effect of gender on wages
  - `matchingdata.csv`, data from South Africa's LFS
\vfill

```{r match1, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
sums <- df %>%
        group_by(female) %>%
        summarise(meanwage = mean(hourlyincome), 
                  agemployment = mean(agemp),
                  african = mean(groupafrican), 
                  coloured = mean(groupcoloured),
                  married = mean(married),
                  age = mean(age),
                  noschooling = mean(no_schooling),
                  secondary = mean(secondary),
                  hhchildsuppgrant = mean(hhchildsuppgrant),
                  urban = mean(urb))
sumst <- as_vector(sums[1,])
sumst <- as.matrix(sumst, col = 1)
sumst <- cbind(sumst, as_vector(sums[2,]))
sumst <- sumst[-1,]
colnames(sumst) <- c("male", "female")
kable(sumst,
          booktabs = TRUE, align = "cc", linesep = "", escape = FALSE, row.names = TRUE, digits = 3) %>%
          kable_classic_2()       
```






## Exact matching
\vfill
- One option is to match exactly on observables
  - For example, we could match on age, education, and marital status
\vfill
- This is straightforward with just a couple variables
\vfill






## Exact matching
```{r match2, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
match <- matchit(female ~ married + secondary, data = df, exact = c("married", "secondary"), replace = TRUE)  
summary(match)
```






## Propensity scores
\vfill
- This becomes problematic when we have *many* covariates
  - It is often impossible to find exact matches across all covariates
\vfill
- Enter: the propensity score
  - The propensity score is the probability of being treated, conditional on covariates
  - The key: we can match on the propensity score, rather than on the covariates themselves
\vfill
- Rosenbaum and Rubin (1983), Dehejia and Wahba (2002)
\vfill






## Propensity scores
```{r match3, echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
match <- matchit(female ~ married + secondary + age + hhchildsuppgrant + no_schooling + groupafrican + groupcoloured + agemp, 
                data = df, replace = TRUE, method = "nearest")  
```






## Propensity scores - All observations
```{r match4, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
match <- summary(matchit(female ~ married + secondary + age + hhchildsuppgrant + no_schooling + groupafrican + groupcoloured + agemp, 
                data = df, replace = TRUE, method = "nearest"))
match$sum.all[,1:3]
```






## Propensity scores - Matched observations
```{r match5, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
match <- summary(matchit(female ~ married + secondary + age + hhchildsuppgrant + no_schooling + groupafrican + groupcoloured + agemp, 
                data = df, replace = TRUE, method = "nearest"))
match$sum.matched[,1:3]
```






## Plotting
```{r match6, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
match <- summary(matchit(female ~ married + secondary + age + hhchildsuppgrant + no_schooling + groupafrican + groupcoloured + agemp, 
                data = df, replace = TRUE, method = "nearest"))
plot(match)
```






## Using the propensity scores
\vfill
- You can then use the propensity score in different ways
\vfill
- Examples:
  - Can match treatment units to control units based on propensity scores
  - Can control for the propensity score in a regression
  - Can match and then use something like differences-in-differences
\vfill
- We won't go into detail, as this is just something I want you to *know* about
\vfill






## Checking common support
```{r match7, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, size = "tiny", out.width = "55%", fig.align = "center"}
match <- matchit(female ~ married + secondary + age + hhchildsuppgrant + no_schooling + groupafrican + groupcoloured + agemp, 
                data = df, method = "nearest")
plot(match, type = "jitter", interactive = FALSE)
```




# In-class practice

## In-class practice
\vfill
- Let's now do some in-class practice
\vfill
- Goals:
  - Practice regression
  - Practice creating tables
\vfill
- Data: deserrannoetal.dta (on GitHub)
\vfill
 















